---
title: "ETC5521 Assignment 1"
subtitle: "Tour de france"
team: Grevillea
author:
  - Samuel Lyubic
  - Brendi Ang
date: "`r Sys.Date()`"
output:
    bookdown::html_document2: default
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

# Introduction 

The Tour de France is a cycling tournament that is held in France annually with the winner being the rider with lowest elapsed time after the completion of every stage which spans across 21 stages over 23 gruelling days. It is considered one of, if not, the most prestigious race that elite cyclists can partake in given the level of difficulty of riding through the ranging beautiful French landscapes @{EEB}. Despite being an individual event, riders travel through the whole country going through the scenic French countryside in teams of 8 in order to strategically position the team leader in the best position to win the general classifications. Teams work with their rider to reduce the physical strain of the course, strategically target particular parts and block certain riders from advancing among many other coordinated efforts for the overall benefit of their lead rider @{EngberD}. The teams would pass through long country side roads, steep alpine regions and tight city areas, covering terrain ranging from rolling hills, long flat grounds and steep mountainous incline in their quest for victory. The types of terrain are dispersed across the different stages over the 23 days with the breakdown usually consisting of flat stages, mountain stages, hilly stages and time trial stages. The division of the stages and the combination of terrains across the Tour are one of the reasons the Tour is considered so gruelling and one of the greatest difficulties for riders and teams to overcome as each type of terrain usually fits a rider with specific compositions however, in order to be victorious they must be able to perform across all the terrains and plan their strategy to control the race effectively @{RickmanC}. Given the chess-like strategy that runs tandem with the elite physical attributes and skills required to win the general classification, this report will do a deep analysis into the Tour de France historical data to look at the past results, riders, teams and potential winning blueprint to answer the primary question: _“What are the most important characteristics and attributes to win the tour and how have they evolved over time?”_ with a number of follow up questions that will inspect the range of variables associated with winning the Tour de France in oder to ultimately explain what it takes to come out victorious.
  
## History of Tour De France

Tour De France ("Le Tour") is an annual men's bicycle race that in modern days encompasses a 21-day course that covers approximately 3,500km. It is predominantly held in France while often passing through other countries. Cyclists from around the world gather for a chance to win the prestigious 'Tour De France' trophy, along with a cash prize of €450,000.

Split into various stages:
- flat stages
- hilly stages
- mountain stages (incl. 5 summit finishes)
- individual ad team time trials
- rest days

One stage is performed every day, covering approximately 225km& about 5.5hours to complete. Each stage has a winner, & the rider that completes the most stages in the shortest amount of time wins the overall title. 

## The Jersey Hierarchy

The colour and patterns of jerseys  to look out for in every race. As a brief overview,
the same coloured jerseys are worn within teams, current road champions get to wear their team jerseys with their country's colour& reigning world champions wears their team jerseys with stripes. The jersey hierarchy, which is based on the how old the competition has been implemented, serves as a symbolic purpose:

**Yellow: General classification - Race leader**: "General Classification" is the oldest and main competition sought after by competitors, where the winner of this will win the overall title. after each stage, the rider with the lowest aggregate time gains the privilege to wear the most coveted jersey in the race, the yellow jersey.  

**Red& white polka-dot: Mountains classification - meilleur grimpeur (Best climber)**: With stages containing climbs, points are awarded to riders who reach the summit first. Thus, this worn by the rider who at the start of each stage, has the largest number of climbing points.

**Green: Points classification - Top sprinter**:   
- Worn by rider who has the greatest number of points at the start of each stage, where points are given to first 15 riders to finish a stage. Furthermore, additional points are also given to first 15 riders to cross a pre-determined 'sprint' point in each stage. These jerseys are usually worn by 'sprinters'.

**White: Young rider classification - Most impressive rider, under 26 years old**: Determined the same as general classification, where it is restricted to riders that are under 26 years old.  

## Teams

You may ask, why have teams when there is only one individual champion for the General Classification? '*Domestiques*' (french in servant), a term frequently used to describe riders who support a team leader to win stages and accumulate the least elapsed time at the end of the race. For instance, to set up a rider for a sprint, one rider rides at a high speed, with the team's sprinter following behind to cut wind resistance and conserve energy; This is known as 'Lead-out Trains'. Teams typically consists of 8 riders and are denoted by numbers 1-9, where the team leader wears the lowest number.

# Data description

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')

# load libraries
library(tidyverse)
library(lubridate)
library(kableExtra)
library(tdf)
# For web scraping
library(polite)
library(rvest)
```

```{r read-data, include = FALSE}
stage_clean <- read.csv(here::here("data/stage_data.csv"))
winners <- read.csv(here::here("data/tdf_winners.csv"))
tdf_stages <- read.csv(here::here("data/tdf_stages.csv"))
```

The data used for this analysis originates from the `tdf` R package, which is essentially a container for the  `editions` data frame. This data set is hosted on [GitHub](https://github.com/alastairrushworth/tdf) and is readily available. The data set encompasses the historic data about Tour de France riders, stages and winners.

`winners` corresponds to a condensed version of the `editions` data frame; Each row corresponds to an edition of the Tour De France event for a given year, spanning all the way to the first edition in 1903 to the latest tournament as of today (2019). Some of the variables include:


```{r}
glimpse(tdf_winners)
```

- `winners ` variables contains the overall winner's biographical data and information about the race for a given edition, which comprises of:  
- `start date` start of the Tour De France edition
- `winner name` name of the overall winner (*i.e. winner of the General Classification)
- `distance` aggregate route distance (in km) covered by the entire race
- `time margin` difference between winning time and runner up (in hours) 
- `stage_wins` stages won by the overall winner
- `stages led` number of stages spent in the yellow jersey
- `height` overall winner's height (in meters)
- `weight` overall winner's weight (in kilograms)
- `born` overall winner's date of birth
- `full_name` overall winner's full name
- `nationality` overall winner's nationality


The `editions` dataset contains variable `stage results`, which represents a list of lists where each element contains a list of stage results pertaining to a particular year or edition of Tour De France. When we unnest, or flatten it back to its regular columns, we get `tdf_stages` and `stage_clean` dataset.


```{r explore-tdf-stages, include = FALSE}
glimpse(tdf_stages)
```

The `tdf_stages` data set depicts the details of the stage winners at each stage of an edition, with stage type/terrain, the distance between the start location and end location of the stage. Some of the variables includes:
- `stage`: Stage number of the edition
- `date`: Date of stage
- `origin` and `destination`: Start and end location of the stage
- `distance`: Distance (in KM) for the stage
- `type`: type of stage
- `winner`: Name of the stage winner

```{r explore-stage-data, include = FALSE}
glimpse(stage_clean)
```

The `stage_clean` is expands on the `tdf_stages` dataset, including results of all riders and their ranks at every stage where each row corresponds to the result of a particular rider such as points accumulated for the stage. This data set includes variables such as:

`edition`: Race edition pertaining to the year
`rider`: The name of the rider/participant
`age`: Age of rider at that point in time
`rank`: Rank of the rider for the stage
`elapsed`: Time elapsed for the stage


## Problems encountered 

Initially, we downloaded all the data set (.csv format) directly from the ['tidytuesday github link'](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-04-07/readme.md), which was the most convenient way to load and read-in the Tour De France data set. However, we discovered errors in the `time` and `elapsed` variables in the `stage_clean` where the period objects were incorrect. Thus, we opted to use the cleaning script found in the same github link and made some adjustments such as changing the object names to reproduce the same data set that was intended. 

Furthermore, `tdf_stages` data set only provided data till the 2017 edition while the actual dataset covered all the way to 2019. Accordingly, we web scraped the data from these the Wikipedia (for [2018](https://en.wikipedia.org/wiki/2018_Tour_de_France) and [2019](https://en.wikipedia.org/wiki/2019_Tour_de_France)) to obtain the stages data set for 2018 and 2019. These Wiki pages corresponds to how the actual data set obtained its data, making it straightforward to structure the data to a tidy format analogous with the `tdf_stages` data set, assuring smooth binding of the data.


# Analysis and findings


## Time and place for speed

The following section will be analysing the overall winners speed and how it changes from stage to stage relative to the rest of the field in order to understand when winning riders believe its most vital to increase their pace over their competitors and increase the elapsed time gap between them. In order to conduct this analysis the rider average speeds in km/h were calculated for each terrain type across every stage. The calculation was done by dividing the elapsed time over the total distance variables from the _stage_clean_ and _tdf_stages_ dataset's. The following analysis will be focusing on the Tour's since 1988 as this is known as the begining of the EPO era for relativitey, given the extreme performance enhancing nature of doping that became prominent from this point onwatds due to the difficulty surrounding its detection @{McKay2013}.

```{r plot-speeds, fig.width = 14, fig.cap = "Average stage speeds across the three main terrains for overall winners in km/h, since 1988"}
#speeds across different stages
winner_stages_combined %>%
  ggplot(aes(x = as.factor(stage_number),
             y = avg_speed,
             color = Type_terrain)) +
  geom_point() +
  geom_point(data = rest_field_stages_combined,
             aes(x = as_factor(stage_number),
                 y = avg_speed,
                 color = the_field,
                 alpha = .4)) +
  theme_linedraw() +
  ggtitle("Overall winners average stage speeds") +
  ylab("Average Speed, km/h") +
  xlab("Stage Number") +
  scale_y_continuous(breaks = seq(0, 50, 5)) +
  labs(color = "Stage Types") +
  facet_wrap(~Type_terrain)

```

Figure \@ref(fig:plot-speeds) illustrates the average stage speed for all overall winning riders since 1988, across all the stages and stage types with stage number on the x axis and the average speed in kilometers per hour on the y axis. _Flat_, _Hilly Stage_, _Mountain_ and _Time Trial_ represent the speeds of the overall winners for each of those terrains and _The Field_ represents the rest of the competitors. From the figure it is evident:

*Flat Stages*
    - The overall average pace from the winners and the field stays relatively consistent for the duration of the course, with speeds increasing slightly towards the middle and end of the Tour, until the legacy "ChampsEssay" rounds theom 20-22. 
    - Overall winners average speed are relatively equal with the average speed of the field for the first 8 stages, with overall winners being 0.3km/hr faster over the first two and 0.34km/h faster in stage 9, which registers the maximum speed for the tour of 43.95km/h for overall winners and 43.56km/h for the fied. 
    - The fild eclipsing the overall winners in stage 10 by 0.43km/h. 
    - The combined speeds show a decreasing trend from stage 12 to stage 14, with overall winners having their largest speed difference  of 0.96km/h, 0.63km/h and 0.96km/h across those three stages. The field produces marginally higher speeds then the overall winners across if 15 to 17. The overall winners show at stage 18 they make the largest difference in average pace with a speed of 43.64km/h which is 1.3km/h greater then the fields average speed of 42.34km/h for the stage.
  - The figure would indicate that across flat stages the field and winners are relatively equal however, when the flat types fall on stage 12, 13, 14 and 19 overall winners seem to look to create the largest gap between them and the field and reduce their elapsed time along the flat terrain.

*Mountain stage types*
    - The combined average speed for both overall winners and the field have a slightly increasing trend from rounds 2 to 5 and then are take a declining trend for the rest of the tour with a few up ticks in speed from round 6 to 8 and from 11 to 13. The maximum speed for over all winners is 41.85k/h and 40.56km/h for the field, which is recorded in stage 5. 
    - Overall winners show to eclipse the fields averge pace at every mountain stage of the Tour. The largest gaps are present across the middle stages from 6 to 10, 12 to 15 and 17 to 20 with the largest difference being 2.04km/h in stage 9, 1.99km/h in stage 12, 2.13km/h in stage 13 and the largest difference in stage 20 of 2.18km/h
  - The figure indicates that when mountain terrain type's fall on stages 6 to 10, 12 to 15 and 17 to 20,  overall winners look to attack and create greater elapsed time seperation from the field at these points. All these stage fall along the declining pattern of the overall speed trend of the mountain stages, indicating that the overall winners target harder stages to impose their presence and ability to create the largest gap between their pace and the field in order to try and and create a larger seperation with their elapsed time.

*Hilly Stages*
    - The combined pace shows similar variation for both overall winnres and field across the first three rounds. This is followed by four different periods of declining speeds trends whereby combined speed declines from stage 3 to stage 7 and then increases at stage 8 and then average speeds decline to stage 12, with another increase at 13 and decline to 15 for overall winners, from there they increase in speed, and further decline for the field followed by relatively similar variation for both speeds until stage 20.
    - Overall winners show to have greater average speed then the field the whole way through the Tour except for stages 10, 17 and 20 where the overall winners speed is marginally eclipsed by the field. The greatest average speed difference come when hilly states fall on 8, 12, 13, 16 and 19 which are 2.36km/h, 1.80km/h, 2.06km/h, 2.33km/h and 2.07km/h, which indicates overall winning riders look to maximise their time difference across these stages when they are hilly types. 
  
*Time Trial*
    - Time trial combined average speed follows a relatively declining trend through to round 16 where there is a the lowest average speed since 1988 recorded of 22.53km/hr for overall winners and 20.31km/hr for the field. Speeds increase through to stage 18 and level out there.
    - The time trial average speeds show that for time trials, the overall winners have the largest speed gap between them and the field, for all stage types. Overall winners hold larger gap for the longest stretch of stages relative to the the differences present in the other three stage types. 
    - This would indicate that overall winners are far more elite then the field at time trials and look likley to minimise their race time along them no matter what stage of the tour the time trial stage falls on. The maximum speed of across the time trials for the overall winners and the field cames when the time trial falls on stage one with speeds of 49,53km/h and 48.13km/h respectively. 

## Where does the greatest difficulty lie?

The following section will be analysing the incompletion rate across the different types of stages. Each stage carries with it it's own unique difficulties and understanding where riders have historically shown the greatest weakness could be benefical at understanding stages that should be targetted and where energy needs to be conserved for. In order to conduct this analysis the total number riders over who failed to complete each stage was divided by the total number competitors who have competed in the stage over time. The _stage_clean_ dataset lists riders who "did not finish", "over the time limit" and "not qualified" as "DNF", "OTL" and "NQ" as their rank for the stage, these ranks were tallied up for every stage as well as the particpants for every stage and then divided over each other to produce the percentage rate of incompletion for each stage.  


```{r DNF-OTL-NQ, fig.cap = "The percentage of riders that have not finished the main for stages, since 1969"}
total_riders_dnf_terrain %>%
  ggplot(aes(x = fct_reorder(Type_terrain, -pct_not_finish),
             y = pct_not_finish)) +
  geom_col(stat = "identity") +
  ggtitle("The percentage of riders that failed to finish") +
  xlab("Stage Type") +
  ylab("Percentage") +
  theme_linedraw() 
```

Figure \@ref(fig:DNF-OTL-NQ) visualises the percentage of total riders that have failed to complete the specific stage types with the stage type on the x axis and the percentage on the y axis. The Mountain and Hilly stages show to be the hardest terrains for riders to finish. Percentage of riders who have failed to complete each stage are:

  - Mountain stage: 2.13% of all riders that have entered the stage have failed to complete it
  - Hilly stage: 1.61% of all riders that have entered the stage have failed to complete it
  - Flat stage: 0.61% of all riders that have entered the stage have failed to complete it
  - Time trial: 0.29% of all riders that have entered the stage have failed to complete it

Time trial's have the least percentage of failed to complete riders with .29% and Flat stages being over double that. Time trial stages are run individually and are usually shorter distances on most often on more flatter terrain these characteristics can make it safer with reduced physical risk and collision risk. This may show that time trials are vital stage to master given their low incompletion rate which could be due to their safer conditions, which could allow for skilled riders to maximise their abilities and use these stages to improve their time gap with competitors. Furthemore, given mountain stages and hilly stages show higher incompletion rates, it would indicate that tailoring training and strategy to these stages could be benefical in the outcome of a riders race as well as, given the greater historical difficulty relative to the outher stages as shown by the higher rate, these could be stages where a rider looks to target the competition by making these stages a strength in order to increase their position when other riders may struggle more. 

## Are elite riders similar?

The following section will be analysing the different body compositions of overall winners compared with the elite stage riders, to assess if their are similarites or differences among the different elite rider styles and what stage type body type may be best suited for overall victory. The stage type body types have been broken down into climber for elite mountain stage riders, sprinter for elite flat ground stage riders and individual time trial for elite indvidual time trial riders. The elite stage riders were identified by agregating the individual stage winners and identifying the riders with most wins for each stage type. The riders were selected from where the largest gap fell to the next rider, for climbers and time trialist's this included six selection and 5 for sprinters however, for overall winners the upper bunch of multiple time winners only consisted of 3 observations so all the winners since 1988 were shown. The physiological data was not provided by the dataset being used in order to conduct the analysis each riders height and weight details were sourced from cyclingranking.com @{CycleRank} and input into tibbles and then combined for the analysis to take place. 

```{r body-types-combine-and-plot, fig.cap = "Comparison of overall winner, sprinter and climbers body types"}
rider_body_types %>%
  ggplot(aes(x = weight,
             y = height)) +
  geom_point() +
  facet_wrap(~rider_type) +
  theme_linedraw() +
  ggtitle("Comparison of body types") +
  ylab("Height, m") +
  xlab("Weight, kg") 
ggplotly()
```

Figure \@ref(fig:body-types-combine-and-plot) visualises the body types of climbers, time trialist, sprinters and the overall winner with weight in kg along the x axis and height in metres on the y axis. The figure illustrates that elite individual time trialist's and sprinters have relatively similar body compositions with most riders falling in the upper right quadrant. Elite time trialists show most of their observations between 75kg and 76kg wwith a height between 1.83m and 1.86m, while sprinters show most of their observations between the heigh of 1.83m and 1.84m and 75 to 80g. Climbers and overall winners take a somwhat similar cone shape, with a relatively similar spread of weight however, climbers tend to be shorter then some overall winners. Overall winners mostly weighted between 65 and 75kgs with their height varying between 1.72m and 1.90

Figure \@ref(fig:body-types-combine-and-plot) shows that climbers have the most similar body type to over all winners which could indicate that have a lower weight could be more beneficial for general classification victory at the Tour, with height not showing to overly impact results according to this figure.

## Which way is age going?

Age often plays a big factor in the assessment of an athelete potential performance. With the developments in the medical, technological and scientific fields athelete's are seeming to continue to compete for longer. The following section will be analysing the age of all the competitors in the Tour de France and how it has changed over time. The aveage age for competitors was collectively calculated for each year starting from 1988. 

```{r median-mean-age, fig.cap = "The averge age of riders at each Tour de France"}
mean_age_tdf <- stage_clean %>%
  filter(year >= "1988") %>%
  select(year, rider, age) %>%
  group_by(year, rider) %>%
  arrange(rider) %>%
  group_by(year, rider) %>%
  slice(1) %>%
  group_by(year) %>%
  summarise(mean_age = mean(age, na.rm = TRUE)) %>%
  arrange(desc(year))

mean_age_tdf %>%
  ggplot(aes(x = year,
             y = mean_age)) +
  geom_point() +
  geom_smooth(alpha = .1, se = FALSE, color = "#FFBE33") +
  theme_linedraw() +
  ggtitle("Average rider age") +
  xlab("Year") +
  ylab("Average age") +
  scale_x_continuous(breaks = seq(1990, 2020, 5))
```

Figure \@ref(fig:median-mean-age) visualises the average ages of contesants from 1988 to 2020 with the year on the x axis and the average age on the y axis. The figure shows the average age has been increasing over time going from an average age of 27.1 in 1998 to a peak of 29.67 in 2012 and ending up with an average age of 29.22 at the 2019 Tour. The advancement of cycling technolgies and development of equipment such as lighter frames and more aerodynmaic attire @{Haake2009} in conjunction with improvements in best practice and integration of sports science, could be a reason behind riders extending their career as physical ailments that would have halted riders in the past are now being dealt with more effectively and proactively @{SmithM} thus allowing riders to extend their careers.

### How has age impacted overall results?

```{r age-winner, fig.cap = "Age of overall winners"}
winners %>%
  filter(start_date >= "1988-07-02") %>%
  mutate(year = year(start_date)) %>%
  ggplot(aes(x = year,
                 y = age)) + 
  geom_point(point = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "#FFBE33") +
  theme_linedraw() +
  ggtitle("Age of overall winners") +
  ylab("Age") +
  xlab("Date") +
  scale_x_continuous(breaks = seq(1988, 2019, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Figure \@ref(fig:age-winner) visualises the age of the overall winner since 1988 with the date on the x axis and the age on the y axis. A slight upward trend is present with quite a bit of variability in amongst the time frame which indicate that the age of winning riders is increasing however, it does not seem to be the most critical element given the distribution on both sides of the trend line. It is also worth noting that, given the EPO era that this time frame is inspecting, the increased riding/winng age could also be influenced by illicit substance use, as was notoriously the case with Lance Armstrong @{WilsonJ} who won seven straight tours from 1999 aged 27 through to 2005 aged 32 as well as less well known riders such as Bjarne Liis who won 1996 aged 32 and also admitted to doping @{FarrandS} as well as Miguel Indurain who won 5 tours in a row from 1991 to 1995 @{FotheringhamA}. These substance based results are the ones that are known and there could potential be others which may have influenced the dataset and skewed the results. 

Although Figure \@ref(fig:age-winner) shows a slight upward trend in winning age, the three diagonal lines present are related to three winners who won consecectuvely with two of them being flagged for doping and the use of the performance enhancers is most likely the reason why they were able to win one, let alone that many in a row thus skewing the age trend upward fallaciously. Overall, this would indicate that being younger or older may not play the largest factor in the chance to claim victory at the Tour de France.  

## Is being part of team important?

Given the strategic nature and benefits that riders gain from their team, the following section will be analysing if being part of a particular team may make a rider more likely to win the Tour given how prevelant that team produces a winner. In order to assess the most prevelant teams in Tour de France, the data set was filtered to show results from 1969 which saw the re-introduction and professional sponsorship of teams @{McGann}. 

```{r most-prevelant-teams, fig.cap "Displays the teams with the most wins since 1969"}
team_names_to_date %>%
  select(start_date, winner_name, winner_team) %>%
  group_by(winner_team) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n >= 3) %>%
  ggplot(aes(x = fct_reorder(winner_team, n),
             y = n)) +
  geom_col(stat = "identity") +
  theme_bw() +
  ggtitle("Teams with the most wins") +
  xlab("Team Name") +
  ylab("Number of wins") +
  scale_y_continuous(breaks = seq(0, 8, 1))
```

Figure \@ref(fig:most-prevelant-teams) displays the most prevalent teams in Tour history since 1969 with the teams on the x axis and the number of wins on the y axis. This figure shows 5 teams which have domianted the results and collectively represented 60% of the winners since 1969. Specifically, Team Ineos who are considered the most dominant team in Tour history credit their performance to improved training habits and implementation of recovery techniques, evolving technologies and marginal gains across the board which collectievly add up to an overall better performance across the team. Furthermore, the financial backing of these larger teams allwo them to attract top tier riders, coaches and staff given the pay packets they are able to provide as well as being able to afford all the the state of the art practices that assist performance thus potentially improving their dominance relative to smaller teams @{LindseyJ}. This may indicat that a rider may have a better chance of winning the Tour de France depending on the team they choose/are chosen for.

## Does it matter where you're from?

Different countries have different characteristics and elements that shape atheletes in different ways, the following section will look to analyse if particular countries are producing more winners and how the country a winner is from has changed over time.

```{r winner-nationality, fig.cap = "Figure displaying a time series of the countries each winner is from"}
winners %>%
  mutate(year = year(start_date)) %>%
  ggplot(aes(x = year,
             y = nationality)) +
  geom_point() +
  theme_linedraw() +
  ggtitle("Nationality of overall winner") +
  xlab("Year") + 
  ylab("Nationality") +
  scale_x_continuous(breaks = seq(1900, 2020, 5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
```

Figure \@ref(fig:winner-nationality) visualises the distribution of countries each winner has been from with the year of the Tour on the x axis and the nationality on the y axis. The figure shows that Up until 1985 France and Belgium were the two most dominant countries from which winning riders were from. In 1986 the U.S. saw their first win and the most dominant nations since then have been the U.S. Spain and Great Britain. This could be a result of globilisation of the sport and access and mobility with regards to training and travel, making it more accessible for riders, coaches and teams to train and operate in the most desired regions with the highest calibre of trainers. However, it should be noted this change comes around the time of the "doping era" which clearly has a large factor in the wins tallied by the United States. Given the freedom of movement and access to terrains and training being accessbile by most athletes, it would still seem a few countries dominate the winning colomn each year which could indicate where the best training partners and highest level of performance and training takes place.

```{r, eval = FALSE}
ggplot(winners ) +
  geom_bar(aes(x = year,          
               y = distance),
           stat = "identity") +
  theme_classic() 
```

```{r}
# insert year variable
winners  <- winners  %>%
  mutate(start_date = as.Date(start_date)) %>%
  mutate(year = year(start_date))
```


```{r tdf_dist}
tdf_dist <- ggplot(data = winners ) +
  geom_point(aes(x = year,
                 y = distance),
             colour = "#ff4d4d") +
  xlab("Year") +
  theme_classic()
```

```{r tdf-avg-speed}
tdf_avg_speed <- ggplot(winners ) +
  geom_point(aes(x = year,
                 y = distance / time_overall),
             colour = "blue") +
  labs(y = "Average Speed",
       x = "Year") +
  theme_classic()
```

```{r combine-dist-avg-speed}
gridExtra::grid.arrange(tdf_dist, tdf_avg_speed, ncol = 2)
```

- The total distance covered are decreasing over the years& has been fluctuating around 3,500km over the past two decades
- At first glance, it may seem that riders have it easy with the reduced distance.
- However, the average speed for each rider had been increasing gradually over the years
- This may be due to better professionalism in the sport, equipment& gears and smarter training.


```{r}
ggplot(winners) +
  geom_point(aes(x = year,
                 y = nationality,
                 colour = nationality),
             alpha = 0.6) +
  theme_minimal() 
```

- Many emerging countries such as Great Britian
- Countries that dominated in the past (e.g. France) are not winning as much now
# Quest for the Yellow Jersey

## Procedure
  
The purpose of this section is to explore the trends of overall winners. We start by answering the most basic question: "Do overall winners win all stages through the edition?". Here, we discovered that what some overall winners achieved in the early decades have not been achieved in contemporary editions. For instance, wearing the yellow jersey for all stages in the edition. So, we decided to hone in on the last two decades, 2000 - 2020. We studied how overall winners were ranked in each stage and stage type to examine if these riders specialised in any particular domain such as sprinting or climbing. For this section, we utilised regular expressions to manipulate character strings and relational joins to join data sets with matching observations.


```{r wrangle-stage-data, warning = FALSE}
# Parse stage number
stage_clean <- stage_clean %>%
  mutate(stage = str_extract(stage_results_id, "\\d+")) %>% 
  mutate(stage = as.factor(stage)) # change to factor for joins

# parse elapsed into lubridate::period variable
stage_clean <- stage_clean %>%
  mutate(elapsed = hms(elapsed))
```

```{r tdf-stages-2000s, warning = FALSE}
# Add year variable for joining table
tdf_stages <- tdf_stages %>% 
  mutate(date = as.Date(date)) %>%
  mutate(year = year(date))

tdf_stages_2000s <- tdf_stages %>%
  filter(between(year, 2000, 2019)) %>%
  rename(stage_winner = winner,
         stage_winner_country = winner_country)

# Change "P" to stage 0
## Represents prologue of the race, usually determines yellow jersey for race 1
tdf_stages_2000s$stage <- ifelse(tdf_stages_2000s$stage == "P", "0", as.factor(tdf_stages_2000s$stage)) 

# include year variable for join
winners <- winners %>%
  mutate(year = year(start_date))

# Extract overall winners' names
winners_2000s <- winners %>%
  filter(between(year, 2000, 2019)) %>%
  select(winner_name, year)

# Change order of winner names to match with stage_clean rider names
winners_2000s$winner_name <-
  str_replace(string = winners_2000s$winner_name, pattern = "([^ ]+) ([^ ]+)", replacement = "\\2 \\1")

# Extract stages details 
stage_winners_2000s <- stage_clean %>%
  filter(between(year, 2000, 2019)) %>%
  # include stage data of overall winner only
  inner_join(winners_2000s, by = c("year", "rider" = "winner_name")) %>% 
  left_join(tdf_stages_2000s, by = c("year", "stage")) %>% # Add stage details
  mutate(rank = as.numeric(rank)) # Change rank to numeric for plotting

# Filter out one NA rank
stage_winners_2000s <- stage_winners_2000s %>%
  filter(!is.na(rank)) %>%
  # Add decade variable, for plotting
  mutate(decade = case_when(
           between(year, 2000, 2009) ~ "2000 - 2009",
           between(year, 2010, 2020) ~ "2010 - 2019")) %>%
  mutate(year = as.factor(year))

stage_winners_2000s$stage_winner <- str_replace(string = stage_winners_2000s$stage_winner, pattern = "\\[.+\\]", replacement = "")

stage_winners_2000s <- stage_winners_2000s %>%
  # remove unnecessary paranthesis
  mutate(stage_winner = str_replace(string = stage_winners_2000s$stage_winner, pattern = "\\[.+\\]", replacement = "")) %>%
  # change sequence of names to match winner names
  mutate(stage_winner = str_replace(string = stage_winners_2000s$stage_winner, pattern = "([^ ]+) ([^ ]+)", replacement = "\\2 \\1"))
```

## A fallacy: Overall Winners leads all stages

```{r stages-led-plot, fig.cap = "How often do Overall Winners Wear the Yellow Jersey"}
ggplot(winners) +
  geom_point(aes(x = year,
                 y = stages_led,
                 ),
           alpha = 0.8) +
  geom_point(data = subset(winners, year %in% c(1928, 1934)), # Riders who led 22 stages
             aes(x = year,
                 y = stages_led,
                 ),
             colour = "#bc3a3a") +
  # Add text on years 1928 and 1934
  geom_text(data =  subset(winners, year %in% 1928), 
            aes(x = year * 0.9977, y = stages_led * 1.003, 
                label = "1928", 
                fontface = "bold")) +
  geom_text(data =  subset(winners, year %in% 1934), 
            aes(x = year * 1.002, y = stages_led * 1.003, 
                label = "1934", 
                fontface = "bold")) +
    geom_line(aes(x = year,
                 y = stages_led
                 ),
              alpha = 0.5) +
  scale_x_continuous(breaks =seq(from = 1900, to = 2020, by = 10),
                     labels =seq(from = 1900, to = 2020, by = 10)) +
  labs(y = "Stages led by Overall Winner",
       x = "Year") +
  theme_minimal() +
  ggtitle("How often do Overall Winners wear the Yellow Jersey?")
```

Many speculate that overall winners are all-rounder cyclists who consistently don the yellow jersey by winning all stages. Nevertheless, winning all 22 stages is not infeasible as Nicolas Frantz and Antonin Magne have both proven it in 1928 and 1934 respectively.

However, in majority of the editions, the line plot (\@ref(fig:stages-led-plot)) invalidates the conjecture, demonstrating that the stage led by overall winners have been shown vary greatly across most editions. The plot further reveals that overall winners can lead as little as 1 stage, indicating that they only led the tour at the very last stage. In addition, overall winners seem to lead in lower number of stages in the recent years as compared to earlier years.

## Ranks of Overall Winners in Different Stages

```{r ranks-by-stage-boxplot, fig.cap = "Ranks of Overall Winners across Stages"}
stage_winners_2000s %>%
  mutate(stage = as.integer(stage)) %>% 
  arrange(desc(stage)) %>%
ggplot(aes(y = as.factor(stage),
           x = rank,
           group = stage)) +
  geom_boxplot() +
  geom_point(alpha = 0.3) +
  theme_minimal() +
  scale_x_continuous(breaks = c(seq(from = 0, to = 150, by = 10))) +
  labs(y = "Stage",
       x = "Rank") +
  ggtitle("Ranks of Overall Winners across Stages from 2000 to 2019") +
  theme(plot.title = element_text(face = "bold", size = 14))
```

The box plot (Figure \@ref(fig:ranks-by-stage-boxplot)) depicts the ranks of al overall winners from 2000 - 2019 for each stage. In most cases, overall winners are ranked lower than 30, which may imply that the winners have a consistent pace that puts them in these ranks. Moreover, it can be observed that in the last few stages, particularly 17-20, overall winners pick up their pace and usually attain top 10 rankings This may suggest that overall winners are wearing the yellow jersey at this point forward. By the same token, this reiterates that the consistent pace of overall winners, where riders who exerted more energy in earlier stages are unable to keep up with the overall winners. 

Why are the ranks of overall winners in stage 21 rather high and peculiar? Stage 21, or better known as the "Champs-Élysées" stage, representing a symbolic street in Paris, France. As a tradition, riders set a truce to their competition to have a celebratory moment of tranquillity, chatting and even celebrating with glasses of champagnes in the final kilometres. For this reason, the rankings of overall winners are diverse.

## Do Overall Winners Thrive in Certain Stage Types or Terrain?

```{r winners-by-stage, fig.width = 8}
# Categorise stages into 4 main stages in tdf
stage_winners_2000s <- stage_winners_2000s %>%
  mutate(type = case_when(
    str_detect(type, "mountain") ~ "Mountain",
    str_detect(type, "Mountain") ~ "Mountain",
    str_detect(type, "Flat") ~ "Flat",
    str_detect(type, "trial") ~ "Time trial",
    str_detect(type, "Hilly") ~ "Hilly")) %>%
  filter(!is.na(type)) # filter out 2 uncategorised stage type 
```

```{r stage-distance-2000s}
# Compute total distance for an edition
stage_winners_2000s <- stage_winners_2000s %>%
  group_by(edition) %>%
  mutate(total_dist = sum(distance, na.rm = TRUE)) # remove one row with NA distance
```

```{r stage-type-prop, fig.cap = "Proportions of stage type for an edition (2000 - 2020)"}
ggplot(stage_winners_2000s) +
  geom_bar(aes(x = year,
               y = total_dist,
               fill = type,
               group = type),
           stat = "identity",
           position = "fill",
           alpha = 0.8) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_brewer(palette = "Dark2") +
  labs(y = "Total Distance",
       x = "Year") +
  ggtitle("Proportions of stage type for an edition (2000 - 2020)") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 14))
```

Despite century steeped traditions, tour organisers arrange the race course in a different manner each year. Nonetheless, it is with a goal in mind, to include a balance of the different types of races. In most editions, the largest proportion of the rest are mountain stages, followed by flat stages (Figure \@ref(fig:stage-type-prop)). In 2007, the edition only consisted of mountain stages and time trials. 


```{r ranks-by-stage-tbl}
# Filter to stages that overall winners won
rank_summ <- stage_winners_2000s %>%
  filter(rank > 5) %>%
  group_by(type) %>%
  tally() %>%
  arrange(desc(n)) 

rank_summ <- stage_winners_2000s %>%
  group_by(type) %>%
  tally() %>%
  inner_join(rank_summ, by = "type") %>%
  rename(total_stage_type = n.x,
         ranked_below_5 = n.y) %>%
  # Find proportion& add '%' symbol
  mutate(prop_rank_type = paste(round((ranked_below_5/total_stage_type)*100, 
                                      digits = 2),
                                "%", sep = "")) %>%
  relocate(total_stage_type, .after = ranked_below_5) %>%
  rename(`Stage Type` = type,
         `Stage Type Total Count` = total_stage_type,
         `Ranked 5th or below` = ranked_below_5,
         `Proportion` = prop_rank_type) 

rank_summ %>%
  kable(caption = "Proportion of Overall Winners Ranked 5th and Below by Stage Type") %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

```{r ranks-by-stage-type, fig.cap = "Rankings of Overall Winners (2000 - 2019) by Stage Type"}
stage_winners_2000s %>%
  ggplot(aes(x = type,
           y = rank)) +
  geom_violin(aes(fill = type),
              width = 1.2,
              alpha = 0.7) +
  geom_boxplot(width = 0.2) +
  geom_point(alpha = 0.2) +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(breaks = c(seq(from = 0, to = 100, by = 10),
                                seq(from = 100, to = 150, by = 25))) +
  theme_bw() +
  labs(x = "Type",
       y = "Rank") +
  ggtitle("Rankings of Overall Winners (2000 - 2019) by Stage Type") +
  theme(legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(face = "bold", size = 14))
```

Figure \@ref(fig:ranks-by-stage-type) illustrates that overall winners are versatile riders who performs relatively well at each stage type. This is also manifest in table \@ref(tab:ranks-by-stage-tbl), which establishes that overall winners are ranked 5 and below in approximately 7 in 10 races for all types of stage. This plot also reinforces the idea that overall winners does not always perform great in all stages, as seen with the variability of ranks at each type of stage.

It is almost always that the overall winners are formidable at climbing mountain. Considering the steep climbs, thin air at high altitudes and narrow descents in mountain stages, these stages have the largest potential for time gaps and for riders to make the grade and set them apart from other riders. Table \@ref(tab:ranks-by-stage-tbl) demonstrates that overall winners thrive in mountain stages, where they are ranked 5 or less in at least three out of four mountain stages. Although this may be true, Figure \@ref(fig:ranks-by-stage-type) also show that one can also expect large variability in rankings of overall winners when it comes to these gruelling mountain stages.

Amongst the different stage types, Figure \@ref(fig:ranks-by-stage-type) shows that the rankings of overall winners had the least median rank and smallest variations when it came to time trials. This implies two things; First, overall winners must be physically skilled to win individual time trials. Individual time trials tests their endurance and solo strengths by forcing them to negotiate the course comprising of curved roads, small-scale slopes and multiple terrain types while keeping a steady and appropriate pace. Next, it conveys that they customarily have great teams as it requires precise strategy and cooperation amongst the team. For instance, team's *domestiques* (French for servants) work in unison to allow for drafting (shield from wind resistance by riders in front) to facilitate a smoother ride for the team leader and ensure he arrives at crucial moments of the race with little energy expended. This is especially complex as teams must find an ideal pace despite their diverse body types and skills. 

With flat stages accounting for nearly one third (`r rank_summ$total_stage_type[1]/nrow(stage_winners_2000s)`) of all stages in a given edition, it is essential for overall winners to be skilled at sprints. Although *Q2* highlighted that flat stages are typically dominated by sprinters, who are usually a muscular rider with a larger body type,  overall winners have proven to be they are successful in these stages too. They are ranked at a median rank of approximately 15 (Figure \@ref(fig:ranks-by-stage-type)) and have attained rank 5 and below in about 73% of the time for all flat stages in the last two decades.


## Did not work 


```{r stage-data-cumsum, eval = FALSE}
# Find total time taken to complete a stage
stage_clean <- stage_clean %>%
  mutate(hour_to_mins = hour(elapsed)*60,
         sec_to_mins = round((second(elapsed)/60), digits = 4),
         mins = minute(elapsed)
         ) 

# Compute total mins per stage for each rider
stage_clean <- stage_clean %>%
  rowwise() %>%
  mutate(total_mins = sum(c(hour_to_mins, sec_to_mins, mins))) # Round to 4 decimal places
  
# Find cumulative sum after each stage 
stage_clean <- stage_clean %>%
  group_by(edition, rider) %>%
  mutate(cum_total_mins = cumsum(total_mins)) 


# Maybe I can colour by the stage type? Need to take stage info from tdf_stages
stage_clean %>%
  filter(year == 2012) %>%
  ggplot() +
  geom_point(aes(x = stage,
                 y = cum_total_mins),
             alpha = 0.01
             ) 

# Create new data set with only winners
# Layer over the graph with another geom_point, coloured in yellow
# Join table using year and rider (because rider may win more than 1 year)
```

```{r test, eval = FALSE}
# Create new data set with only winners
winners <- winners %>%
  mutate(year = year(start_date))

# Create new data set with winners for 2019
winners_2017 <- winners %>%
  filter(year == 2017) 

# Insert stage records by joining table
winners_2017 %>%
  left_join()

# Insert stage records by joining table with regular expression
winner_name <- winners_2017$winner_name
winner_match <- str_c("[", winner_name, "]",  
                      collapse = "|")

str_subset(string = stage_2017$rider, pattern = winner_match)
str_view_all(stage_2017$rider, "Chris Froome", match = TRUE)

# Layer over the graph with another geom_point, coloured in yellow
```



```{r, eval = FALSE}
# Transform factor to original numeric values
stage_clean$stage <- as.numeric(levels(stage_clean$stage))[stage_clean$stage]

# Find first& last stage of each year
finish_summ <- stage_clean %>%
  group_by(year) %>%
  # Filter to stage 1, not 0 because some riders don't race in prologue
  filter(stage == 1 | stage == max(stage)) %>%
  summarise(first_stage = min(stage),
            last_stage = max(stage)) %>%
  # pivot longer to facilitate join
  pivot_longer(cols = first_stage:last_stage,
               names_to = "first_last_stage",
               values_to = "stage")

# Filter to only first& last stages of each edition
stage_finish <- stage_clean %>%
  inner_join(finish_summ, by = c("year", 
                                 "stage")) %>%
  # filter out years with error in rider data
  filter(!year %in% c(1939, 1964, 1966, 1967, 1968, 1972, 1976, 1977)) 

stage_finish <- stage_finish %>%
  group_by(year, stage, first_last_stage) %>%
  summarise(total_riders = n()) 
```


```{r, eval = FALSE}
stage_finish_wide <- stage_finish %>%
  pivot_wider(
    id = year,
    names_from = first_last_stage,
    values_from = c(total_riders, stage)
  ) %>%
  rename(first_stage = stage_first_stage,
         last_stage = stage_last_stage) %>%
  # include % of riders who finished entire race
  mutate(finish_prop = total_riders_last_stage/total_riders_first_stage)
```

```{r, eval = FALSE}
ggplot(data = stage_finish_wide,
       aes(x = year)) +
  geom_point(aes(y = total_riders_first_stage),
             colour = "steelblue") +
  geom_point(aes(y = total_riders_last_stage,
                 colour = total_riders_last_stage),
             colour = "red") +
  geom_segment(
    aes(y = total_riders_first_stage,
        xend = year, 
        yend = total_riders_last_stage), alpha = 0.5) 
```

```{r finish-rate, eval = FALSE}
stage_finish_wide %>%
  ggplot(aes(x = year,
                 y = finish_prop)) + 
  geom_point() +
  labs(y = "% of Riders who Finished all Stages",
       x = "Year") +
  geom_smooth(method = "lm",
              span = 0.2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) + # Remove panel grids
  scale_x_continuous(breaks =seq(from = 1900, to = 2020, by = 10),
                     labels = seq(from = 1900, to = 2020, by = 10)) 
```

# 
