---
title: "ETC5521 Assignment 1"
subtitle: "Tour de france"
team: Grevillea
author:
  - Samuel Lyubic
  - Brendi Ang
date: "`r Sys.Date()`"
output:
    bookdown::html_document2: default
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

# Introduction 

The Tour de France is a cycling tournament that is held in France annually with the winner being the rider with lowest elapsed time after the completion of every stage which spans across 21 stages over 23 gruelling days. It is considered one of, if not, the most prestigious race that elite cyclists can partake in given the level of difficulty of riding through the ranging beautiful French landscapes @{EEB}. Despite being an individual event, riders travel through the whole country going through the scenic French countryside in teams of 8 in order to strategically position the team leader in the best position to win the general classifications. Teams work with their rider to reduce the physical strain of the course, strategically target particular parts and block certain riders from advancing among many other coordinated efforts for the overall benefit of their lead rider @{EngberD}. The teams would pass through long country side roads, steep alpine regions and tight city areas, covering terrain ranging from rolling hills, long flat grounds and steep mountainous incline in their quest for victory. The types of terrain are dispersed across the different stages over the 23 days with the breakdown usually consisting of flat stages, mountain stages, hilly stages and time trial stages. The division of the stages and the combination of terrains across the Tour are one of the reasons the Tour is considered so gruelling and one of the greatest difficulties for riders and teams to overcome as each type of terrain usually fits a rider with specific compositions however, in order to be victorious they must be able to perform across all the terrains and plan their strategy to control the race effectively @{RickmanC}. Given the chess-like strategy that runs tandem with the elite physical attributes and skills required to win the general classification, this report will do a deep analysis into the Tour de France historical data to look at the past results, riders, teams and potential winning blueprint to answer the primary question: _“What are the most important characteristics and attributes to win the tour and how have they evolved over time?”_ with a number of follow up questions that will inspect the range of variables associated with winning the Tour de France in oder to ultimately explain what it takes to come out victorious.
  
## History of Tour De France

Tour De France ("Le Tour") is an annual men's bicycle race that in modern days encompasses a 21-day course that covers approximately 3,500km. It is predominantly held in France while often passing through other countries. Cyclists from around the world gather for a chance to win the prestigious 'Tour De France' trophy, along with a cash prize of €450,000.

Split into various stages:
- flat stages
- hilly stages
- mountain stages (incl. 5 summit finishes)
- individual ad team time trials
- rest days

One stage is performed every day, covering approximately 225km& about 5.5hours to complete. Each stage has a winner, & the rider that completes the most stages in the shortest amount of time wins the overall title. 

## The Jersey Hierarchy

The colour and patterns of jerseys  to look out for in every race. As a brief overview,
the same coloured jerseys are worn within teams, current road champions get to wear their team jerseys with their country's colour& reigning world champions wears their team jerseys with stripes. The jersey hierarchy, which is based on the how old the competition has been implemented, serves as a symbolic purpose:

**Yellow: General classification - Race leader**: "General Classification" is the oldest and main competition sought after by competitors, where the winner of this will win the overall title. after each stage, the rider with the lowest aggregate time gains the privilege to wear the most coveted jersey in the race, the yellow jersey.  

**Red& white polka-dot: Mountains classification - meilleur grimpeur (Best climber)**: With stages containing climbs, points are awarded to riders who reach the summit first. Thus, this worn by the rider who at the start of each stage, has the largest number of climbing points.

**Green: Points classification - Top sprinter**:   
- Worn by rider who has the greatest number of points at the start of each stage, where points are given to first 15 riders to finish a stage. Furthermore, additional points are also given to first 15 riders to cross a pre-determined 'sprint' point in each stage. These jerseys are usually worn by 'sprinters'.

**White: Young rider classification - Most impressive rider, under 26 years old**: Determined the same as general classification, where it is restricted to riders that are under 26 years old.  

## Teams

You may ask, why have teams when there is only one individual champion for the General Classification? '*Domestiques*' (french in servant), a term frequently used to describe riders who support a team leader to win stages and accumulate the least elapsed time at the end of the race. For instance, to set up a rider for a sprint, one rider rides at a high speed, with the team's sprinter following behind to cut wind resistance and conserve energy; This is known as 'Lead-out Trains'. Teams typically consists of 8 riders and are denoted by numbers 1-9, where the team leader wears the lowest number.

# Data description

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')

# load libraries
library(tidyverse)
library(lubridate)
library(kableExtra)
library(tdf)
# For web scraping
library(polite)
library(rvest)
library(plotly)
```

```{r read-data, include = FALSE}
stage_clean <- read.csv(here::here("data/stage_data.csv"))
winners <- read.csv(here::here("data/tdf_winners.csv"))
tdf_stages <- read.csv(here::here("data/tdf_stages.csv"))
```

The data used for this analysis originates from the `tdf` R package, which is essentially a container for the  `editions` data frame. This data set is hosted on [GitHub](https://github.com/alastairrushworth/tdf) and is readily available. The data set encompasses the historic data about Tour de France riders, stages and winners.

`winners` corresponds to a condensed version of the `editions` data frame; Each row corresponds to an edition of the Tour De France event for a given year, spanning all the way to the first edition in 1903 to the latest tournament as of today (2019). Some of the variables include:


```{r}
glimpse(tdf_winners)
```

- `winners ` variables contains the overall winner's biographical data and information about the race for a given edition, which comprises of:  
- `start date` start of the Tour De France edition
- `winner name` name of the overall winner (*i.e. winner of the General Classification)
- `distance` aggregate route distance (in km) covered by the entire race
- `time margin` difference between winning time and runner up (in hours) 
- `stage_wins` stages won by the overall winner
- `stages led` number of stages spent in the yellow jersey
- `height` overall winner's height (in meters)
- `weight` overall winner's weight (in kilograms)
- `born` overall winner's date of birth
- `full_name` overall winner's full name
- `nationality` overall winner's nationality


The `editions` dataset contains variable `stage results`, which represents a list of lists where each element contains a list of stage results pertaining to a particular year or edition of Tour De France. When we unnest, or flatten it back to its regular columns, we get `tdf_stages` and `stage_clean` dataset.


```{r explore-tdf-stages, include = FALSE}
glimpse(tdf_stages)
```

The `tdf_stages` data set depicts the details of the stage winners at each stage of an edition, with stage type/terrain, the distance between the start location and end location of the stage. Some of the variables includes:
- `stage`: Stage number of the edition
- `date`: Date of stage
- `origin` and `destination`: Start and end location of the stage
- `distance`: Distance (in KM) for the stage
- `type`: type of stage
- `winner`: Name of the stage winner

```{r explore-stage-data, include = FALSE}
glimpse(stage_clean)
```

The `stage_clean` is expands on the `tdf_stages` dataset, including results of all riders and their ranks at every stage where each row corresponds to the result of a particular rider such as points accumulated for the stage. This data set includes variables such as:

`edition`: Race edition pertaining to the year
`rider`: The name of the rider/participant
`age`: Age of rider at that point in time
`rank`: Rank of the rider for the stage
`elapsed`: Time elapsed for the stage


## Problems encountered 

Initially, we downloaded all the data set (.csv format) directly from the ['tidytuesday github link'](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-04-07/readme.md), which was the most convenient way to load and read-in the Tour De France data set. However, we discovered errors in the `time` and `elapsed` variables in the `stage_clean` where the period objects were incorrect. Thus, we opted to use the cleaning script found in the same github link and made some adjustments such as changing the object names to reproduce the same data set that was intended. 

Furthermore, `tdf_stages` data set only provided data till the 2017 edition while the actual dataset covered all the way to 2019. Accordingly, we web scraped the data from these the Wikipedia (for [2018](https://en.wikipedia.org/wiki/2018_Tour_de_France) and [2019](https://en.wikipedia.org/wiki/2019_Tour_de_France)) to obtain the stages data set for 2018 and 2019. These Wiki pages corresponds to how the actual data set obtained its data, making it straightforward to structure the data to a tidy format analogous with the `tdf_stages` data set, assuring smooth binding of the data.


# Analysis and findings


## Time and place for speed

The following section will be analysing the overall winners speed and how it changes from stage to stage relative to the rest of the field in order to understand when winning riders believe its most vital to increase their pace over their competitors and increase the elapsed time gap between them. In order to conduct this analysis the rider average speeds in km/h were calculated for each terrain type across every stage. The calculation was done by dividing the elapsed time over the total distance variables from the _stage_clean_ and _tdf_stages_ dataset's. The following analysis will be focusing on the Tour's since 1988 as this is known as the begining of the EPO era for relativitey, given the extreme performance enhancing nature of doping that became prominent from this point onwatds due to the difficulty surrounding its detection @{McKay2013}.

```{r plot-speeds, fig.width = 14, fig.cap = "Average stage speeds across the three main terrains for overall winners in km/h, since 1988"}
#speeds across different stages
winner_stages_combined %>%
  ggplot(aes(x = as.factor(stage_number),
             y = avg_speed,
             color = Type_terrain)) +
  geom_point() +
  geom_point(data = rest_field_stages_combined,
             aes(x = as_factor(stage_number),
                 y = avg_speed,
                 color = the_field,
                 alpha = .4)) +
  theme_linedraw() +
  ggtitle("Overall winners average stage speeds") +
  ylab("Average Speed, km/h") +
  xlab("Stage Number") +
  scale_y_continuous(breaks = seq(0, 50, 5)) +
  labs(color = "Stage Types") +
  facet_wrap(~Type_terrain)
ggplotly()

```

Figure \@ref(fig:plot-speeds) illustrates the average stage speed for all overall winning riders since 1988, across all the stages and stage types with stage number on the x axis and the average speed in kilometers per hour on the y axis. _Flat_, _Hilly Stage_, _Mountain_ and _Time Trial_ represent the speeds of the overall winners for each of those terrains and _The Field_ represents the rest of the competitors. From the figure it is evident:

*Flat Stages*
    - The overall average pace from the winners and the field stays relatively consistent for the duration of the course, with speeds increasing slightly towards the middle and end of the Tour, until the legacy "ChampsEssay" rounds theom 20-22. 
    - Overall winners average speed are relatively equal with the average speed of the field for the first 8 stages, with overall winners being 0.3km/hr faster over the first two and 0.34km/h faster in stage 9, which registers the maximum speed for the tour of 43.95km/h for overall winners and 43.56km/h for the fied. 
    - The fild eclipsing the overall winners in stage 10 by 0.43km/h. 
    - The combined speeds show a decreasing trend from stage 12 to stage 14, with overall winners having their largest speed difference  of 0.96km/h, 0.63km/h and 0.96km/h across those three stages. The field produces marginally higher speeds then the overall winners across if 15 to 17. The overall winners show at stage 18 they make the largest difference in average pace with a speed of 43.64km/h which is 1.3km/h greater then the fields average speed of 42.34km/h for the stage.
  - The figure would indicate that across flat stages the field and winners are relatively equal however, when the flat types fall on stage 12, 13, 14 and 19 overall winners seem to look to create the largest gap between them and the field and reduce their elapsed time along the flat terrain.

*Mountain stage types*
    - The combined average speed for both overall winners and the field have a slightly increasing trend from rounds 2 to 5 and then are take a declining trend for the rest of the tour with a few up ticks in speed from round 6 to 8 and from 11 to 13. The maximum speed for over all winners is 41.85k/h and 40.56km/h for the field, which is recorded in stage 5. 
    - Overall winners show to eclipse the fields averge pace at every mountain stage of the Tour. The largest gaps are present across the middle stages from 6 to 10, 12 to 15 and 17 to 20 with the largest difference being 2.04km/h in stage 9, 1.99km/h in stage 12, 2.13km/h in stage 13 and the largest difference in stage 20 of 2.18km/h
  - The figure indicates that when mountain terrain type's fall on stages 6 to 10, 12 to 15 and 17 to 20,  overall winners look to attack and create greater elapsed time seperation from the field at these points. All these stage fall along the declining pattern of the overall speed trend of the mountain stages, indicating that the overall winners target harder stages to impose their presence and ability to create the largest gap between their pace and the field in order to try and and create a larger seperation with their elapsed time.

*Hilly Stages*
    - The combined pace shows similar variation for both overall winnres and field across the first three rounds. This is followed by four different periods of declining speeds trends whereby combined speed declines from stage 3 to stage 7 and then increases at stage 8 and then average speeds decline to stage 12, with another increase at 13 and decline to 15 for overall winners, from there they increase in speed, and further decline for the field followed by relatively similar variation for both speeds until stage 20.
    - Overall winners show to have greater average speed then the field the whole way through the Tour except for stages 10, 17 and 20 where the overall winners speed is marginally eclipsed by the field. The greatest average speed difference come when hilly states fall on 8, 12, 13, 16 and 19 which are 2.36km/h, 1.80km/h, 2.06km/h, 2.33km/h and 2.07km/h, which indicates overall winning riders look to maximise their time difference across these stages when they are hilly types. 
  
*Time Trial*
    - Time trial combined average speed follows a relatively declining trend through to round 16 where there is a the lowest average speed since 1988 recorded of 22.53km/hr for overall winners and 20.31km/hr for the field. Speeds increase through to stage 18 and level out there.
    - The time trial average speeds show that for time trials, the overall winners have the largest speed gap between them and the field, for all stage types. Overall winners hold larger gap for the longest stretch of stages relative to the the differences present in the other three stage types. 
    - This would indicate that overall winners are far more elite then the field at time trials and look likley to minimise their race time along them no matter what stage of the tour the time trial stage falls on. The maximum speed of across the time trials for the overall winners and the field cames when the time trial falls on stage one with speeds of 49,53km/h and 48.13km/h respectively. 

## Where does the greatest difficulty lie?

The following section will be analysing the incompletion rate across the different types of stages. Each stage carries with it it's own unique difficulties and understanding where riders have historically shown the greatest weakness could be benefical at understanding stages that should be targetted and where energy needs to be conserved for. In order to conduct this analysis the total number riders over who failed to complete each stage was divided by the total number competitors who have competed in the stage over time. The _stage_clean_ dataset lists riders who "did not finish", "over the time limit" and "not qualified" as "DNF", "OTL" and "NQ" as their rank for the stage, these ranks were tallied up for every stage as well as the particpants for every stage and then divided over each other to produce the percentage rate of incompletion for each stage.  


```{r DNF-OTL-NQ, fig.cap = "The percentage of riders that have not finished the main for stages, since 1969"}
total_riders_dnf_terrain %>%
  ggplot(aes(x = fct_reorder(Type_terrain, -pct_not_finish),
             y = pct_not_finish)) +
  geom_col(stat = "identity") +
  ggtitle("The percentage of riders that failed to finish") +
  xlab("Stage Type") +
  ylab("Percentage") +
  theme_linedraw() 
```

Figure \@ref(fig:DNF-OTL-NQ) visualises the percentage of total riders that have failed to complete the specific stage types with the stage type on the x axis and the percentage on the y axis. The Mountain and Hilly stages show to be the hardest terrains for riders to finish. Percentage of riders who have failed to complete each stage are:

  - Mountain stage: 2.13% of all riders that have entered the stage have failed to complete it
  - Hilly stage: 1.61% of all riders that have entered the stage have failed to complete it
  - Flat stage: 0.61% of all riders that have entered the stage have failed to complete it
  - Time trial: 0.29% of all riders that have entered the stage have failed to complete it

Time trial's have the least percentage of failed to complete riders with .29% and Flat stages being over double that. Time trial stages are run individually and are usually shorter distances on most often on more flatter terrain these characteristics can make it safer with reduced physical risk and collision risk. This may show that time trials are vital stage to master given their low incompletion rate which could be due to their safer conditions, which could allow for skilled riders to maximise their abilities and use these stages to improve their time gap with competitors. Furthemore, given mountain stages and hilly stages show higher incompletion rates, it would indicate that tailoring training and strategy to these stages could be benefical in the outcome of a riders race as well as, given the greater historical difficulty relative to the outher stages as shown by the higher rate, these could be stages where a rider looks to target the competition by making these stages a strength in order to increase their position when other riders may struggle more. 

## Are elite riders similar?

The following section will be analysing the different body compositions of overall winners compared with the elite stage riders, to assess if their are similarites or differences among the different elite rider styles and what stage type body type may be best suited for overall victory. The stage type body types have been broken down into climber for elite mountain stage riders, sprinter for elite flat ground stage riders and individual time trial for elite indvidual time trial riders. The elite stage riders were identified by agregating the individual stage winners and identifying the riders with most wins for each stage type. The riders were selected from where the largest gap fell to the next rider, for climbers and time trialist's this included six selection and 5 for sprinters however, for overall winners the upper bunch of multiple time winners only consisted of 3 observations so all the winners since 1988 were shown. The physiological data was not provided by the dataset being used in order to conduct the analysis each riders height and weight details were sourced from cyclingranking.com @{CycleRank} and input into tibbles and then combined for the analysis to take place. 

```{r body-types-combine-and-plot, fig.cap = "Comparison of overall winner, sprinter and climbers body types"}
rider_body_types %>%
  ggplot(aes(x = weight,
             y = height)) +
  geom_point() +
  facet_wrap(~rider_type) +
  theme_linedraw() +
  ggtitle("Comparison of body types") +
  ylab("Height, m") +
  xlab("Weight, kg") 
ggplotly()
```

Figure \@ref(fig:body-types-combine-and-plot) visualises the body types of climbers, time trialist, sprinters and the overall winner with weight in kg along the x axis and height in metres on the y axis. The figure illustrates that elite individual time trialist's and sprinters have relatively similar body compositions with most riders falling in the upper right quadrant. Elite time trialists show most of their observations between 75kg and 76kg wwith a height between 1.83m and 1.86m, while sprinters show most of their observations between the heigh of 1.83m and 1.84m and 75 to 80g. Climbers and overall winners take a somwhat similar cone shape, with a relatively similar spread of weight however, climbers tend to be shorter then some overall winners. Overall winners mostly weighted between 65 and 75kgs with their height varying between 1.72m and 1.90

Figure \@ref(fig:body-types-combine-and-plot) shows that climbers have the most similar body type to over all winners which could indicate that have a lower weight could be more beneficial for general classification victory at the Tour, with height not showing to overly impact results according to this figure.

## Which way is age going?

Age often plays a big factor in the assessment of an athelete potential performance. With the developments in the medical, technological and scientific fields athelete's are seeming to continue to compete for longer. The following section will be analysing the age of all the competitors in the Tour de France and how it has changed over time. The aveage age for competitors was collectively calculated for each year starting from 1988. 

```{r median-mean-age, fig.cap = "The averge age of riders at each Tour de France"}
mean_age_tdf <- stage_clean %>%
  filter(year >= "1988") %>%
  select(year, rider, age) %>%
  group_by(year, rider) %>%
  arrange(rider) %>%
  group_by(year, rider) %>%
  slice(1) %>%
  group_by(year) %>%
  summarise(mean_age = mean(age, na.rm = TRUE)) %>%
  arrange(desc(year))

mean_age_tdf %>%
  ggplot(aes(x = year,
             y = mean_age)) +
  geom_point() +
  geom_smooth(alpha = .1, se = FALSE, color = "#FFBE33") +
  theme_linedraw() +
  ggtitle("Average rider age") +
  xlab("Year") +
  ylab("Average age") +
  scale_x_continuous(breaks = seq(1990, 2020, 5))
```

Figure \@ref(fig:median-mean-age) visualises the average ages of contesants from 1988 to 2020 with the year on the x axis and the average age on the y axis. The figure shows the average age has been increasing over time going from an average age of 27.1 in 1998 to a peak of 29.67 in 2012 and ending up with an average age of 29.22 at the 2019 Tour. The advancement of cycling technolgies and development of equipment such as lighter frames and more aerodynmaic attire @{Haake2009} in conjunction with improvements in best practice and integration of sports science, could be a reason behind riders extending their career as physical ailments that would have halted riders in the past are now being dealt with more effectively and proactively @{SmithM} thus allowing riders to extend their careers.

### How has age impacted overall results?

```{r age-winner, fig.cap = "Age of overall winners"}
winners %>%
  filter(start_date >= "1988-07-02") %>%
  mutate(year = year(start_date)) %>%
  ggplot(aes(x = year,
                 y = age)) + 
  geom_point(point = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "#FFBE33") +
  theme_linedraw() +
  ggtitle("Age of overall winners") +
  ylab("Age") +
  xlab("Date") +
  scale_x_continuous(breaks = seq(1988, 2019, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Figure \@ref(fig:age-winner) visualises the age of the overall winner since 1988 with the date on the x axis and the age on the y axis. A slight upward trend is present with quite a bit of variability in amongst the time frame which indicate that the age of winning riders is increasing however, it does not seem to be the most critical element given the distribution on both sides of the trend line. It is also worth noting that, given the EPO era that this time frame is inspecting, the increased riding/winng age could also be influenced by illicit substance use, as was notoriously the case with Lance Armstrong @{WilsonJ} who won seven straight tours from 1999 aged 27 through to 2005 aged 32 as well as less well known riders such as Bjarne Liis who won 1996 aged 32 and also admitted to doping @{FarrandS} as well as Miguel Indurain who won 5 tours in a row from 1991 to 1995 @{FotheringhamA}. These substance based results are the ones that are known and there could potential be others which may have influenced the dataset and skewed the results. 

Although Figure \@ref(fig:age-winner) shows a slight upward trend in winning age, the three diagonal lines present are related to three winners who won consecectuvely with two of them being flagged for doping and the use of the performance enhancers is most likely the reason why they were able to win one, let alone that many in a row thus skewing the age trend upward fallaciously. Overall, this would indicate that being younger or older may not play the largest factor in the chance to claim victory at the Tour de France.  

## Is being part of team important?

Given the strategic nature and benefits that riders gain from their team, the following section will be analysing if being part of a particular team may make a rider more likely to win the Tour given how prevelant that team produces a winner. The data set will be focusing on winners from 1969 given this was the year that professional teams were re-intriduced into the Tour de France as previously mentioned. In order to assess the most prevelant teams in Tour de France, the data set was filtered to show results from 1969 which saw the re-introduction and professional sponsorship of teams @{McGann}. 

```{r most-prevelant-teams, fig.cap "Displays the teams with the most wins since 1969"}
team_names_to_date %>%
  select(start_date, winner_name, winner_team) %>%
  group_by(winner_team) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n >= 3) %>%
  ggplot(aes(x = fct_reorder(winner_team, n),
             y = n)) +
  geom_col(stat = "identity") +
  theme_bw() +
  ggtitle("Teams with the most wins") +
  xlab("Team Name") +
  ylab("Number of wins") +
  scale_y_continuous(breaks = seq(0, 8, 1))
```

Figure \@ref(fig:most-prevelant-teams) displays the most prevalent teams in Tour history  with the teams on the x axis and the number of wins on the y axis. This figure shows 5 teams which have domianted the results and collectively represented 60% of the winners since 1969. Specifically, Team Ineos who are considered the most dominant team in Tour history credit their performance to improved training habits and implementation of recovery techniques, evolving technologies and marginal gains across the board which collectievly add up to an overall better performance across the team. Furthermore, the financial backing of these larger teams allwo them to attract top tier riders, coaches and staff given the pay packets they are able to provide as well as being able to afford all the the state of the art practices that assist performance thus potentially improving their dominance relative to smaller teams @{LindseyJ}. This may indicate that a rider may have a better chance of winning the Tour de France depending on the team they choose/are chosen for.

## Does it matter where you're from?

Different countries have different characteristics and elements that shape atheletes in different ways, the following section will look to analyse if particular countries are producing more winners and how the country a winner is from has changed over time.

```{r winner-nationality, fig.cap = "Figure displaying a time series of the countries each winner is from"}
winners %>%
  mutate(year = year(start_date)) %>%
  ggplot(aes(x = year,
             y = nationality)) +
  geom_point() +
  theme_linedraw() +
  ggtitle("Nationality of overall winner") +
  xlab("Year") + 
  ylab("Nationality") +
  scale_x_continuous(breaks = seq(1900, 2020, 5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
```

Figure \@ref(fig:winner-nationality) visualises the distribution of countries each winner has been from with the year of the Tour on the x axis and the nationality on the y axis. The figure shows that Up until 1985 France and Belgium were the two most dominant countries from which winning riders were from. In 1986 the U.S. saw their first win and the most dominant nations since then have been the U.S. Spain and Great Britain. This could be a result of globilisation of the sport and access and mobility with regards to training and travel, making it more accessible for riders, coaches and teams to train and operate in the most desired regions with the highest calibre of trainers. However, it should be noted this change comes around the time of the "doping era" which clearly has a large factor in the wins tallied by the United States. Given the freedom of movement and access to terrains and training being accessbile by most athletes, it would still seem a few countries dominate the winning colomn each year which could indicate where the best training partners and highest level of performance and training takes place.

